msqrob2TMT: robust linear mixed models for inferring differential abundant proteins in labelled experiments with arbitrarily complex design

### 2025-06-17 paper/manuscript_tc.tex
Updated latest version upon peer-review

---
title: "msqrob2TMT: robust linear mixed models for inferring differential abundant proteins in labelled experiments with arbitrarily complex design"
authors:
  - name: Stijn Vandenbulcke
    affiliations:
      - ref: UGent
      - ref: VIB
      - ref: Biomed
    equal-contributor: true
  - name: Christophe Vanderaa
    affiliations:
      - ref: UGent
    equal-contributor: true
  - name: Oliver Crook
    affiliations:
      - ref: UK
  - name: Lennart Martens
    affiliations:
      - ref: VIB
      - ref: Biomed
  - name: Lieven Clement
    affiliations:
      - ref: UGent
    email: lieven.clement@ugent.be
    corresponding: true

affiliations:
  - id: UGent
    name: Department of Mathematics, Computer Science and Statistics, Ghent University, Ghent, Belgium
  - id: VIB
    name: CompOmics, VIB Center for Medical Biotechnology, VIB, Ghent, Belgium
  - id: Biomed
    name: Department of Biomolecular Medicine, Faculty of Medicine and Health Sciences, Ghent University, Ghent, Belgium
  - id: UK
    name: Department of Statistics, University of Oxford, Oxford, UK.

filters:
  - ../_extensions/kapsner/authors-block/authors-block.lua

bibliography: references.bib
format:
  pdf:
    number-sections: true
    number-depth: 3
    pdf-engine: latexmk
    pdf-engine-opt: -xelatex
    keep-tex: true
    include-in-header:
      text:
        \usepackage{amsmath, booktabs, caption, longtable, xcolor, ulem}
editor: source
crossref:
  fig-title: Figure     # (default is "Figure")
  fig-prefix: Figure
knitr:
  opts_chunk:
    fig.path: ../figs/
---

@@

```{r, echo =FALSE, message  =FALSE, warning = FALSE, results = 'hide', eval=TRUE}
#classoption: twocolumn
suppressPackageStartupMessages({
    library(tidyverse)
    library(flextable)
    library(gt)
})

dataDir <- "../data/"
plotDir <- "../figs/"
```

@@

High-throughput LC-MS-based proteomic workflows are widely used to quantify differential protein abundance across samples. Relative protein quantification is generally achieved through either label-free or labelled workflows. The latter employ stable isotope labelling techniques such as metabolic and post-metabolic labelling, which gained traction in recent years [@Li2021]. Compared to label-free approaches, labelling offers the advantage of mitigating run-to-run variability in peptide identification and quantification by pooling and analysing multiple samples within a single run. This enables researchers to compare proteomes across multiple conditions or treatments within a single MS run, thereby providing a more comprehensive view of the proteome and enhancing the statistical power of the analysis. Current tandem mass tag (TMT) kits, for instance, support the multiplexing of up to 18 samples in a single run [@Li2021].

However, contemporary experiments often involve increasingly complex designs, where the number of samples exceeds the capacity of a single run. This has far reaching consequences for the downstream data analysis. In labelled experiments with multiple MS runs, biological replicates are often distributed across distinct runs, and additional technical MS runs are frequently included. Similar to label-free approaches, such multi-run experiments are prone to numerous missing peptide intensity values across MS runs [@Brenes2019]. Furthermore, the data of these designs are correlated at multiple levels. Indeed, the MS-technology introduces various sources of technical variation, including run-level effects, channel effects, and spectrum-specific effects, amongst others. Hence, ion intensities from samples multiplexed in the same run are more similar than those from samples multiplexed in different runs. Additionally, quantification does not happen at the protein level but at the level of its identified peptide ions. Peptide ion abundances from the same sample within a run are inherently more similar than those of peptide ions from different samples. At the lowest level, TMT reporter ions from a specific spectrum enable peptide ion quantification across multiple samples, resulting in correlated ion intensities; ion intensities within the same spectrum are more alike than those measured across different spectra. Furthermore, mixtures of multiplexed samples are often run in technical repeats on the MS, introducing yet another level of correlation.  Addressing such a complex correlation structure in labelled experiments necessitates flexible statistical models capable of producing valid tests for differential abundance. However, most proteomics software tools assume independence among the (summarised) ion intensities and can therefore not account for the abovementioned intricate correlation structures in contemporary TMT experiments, which may in turn lead to an inflation of false positives [@Huang2020]. The MSstatsTMT tool [@Huang2020], however, is one of the notable exceptions, as it has been explicitly developed to accommodate experiments with multiple conditions, biological replicate runs, technical repeat runs, and unbalanced designs.

In this publication, we present a bespoke data analysis workflow for labelled MS-based proteomics experiments, embedded in our msqrob2 universe in R/Bioconductor. Initially designed for label-free workflows, msqrob2 has demonstrated advantages over other state-of-the-art tools in terms of both performance and its modular, transparent implementation [@Sticker2020]. In this work, we extend these benefits to the analysis of labelled experiments. Similar to MSstatsTMT, our workflows address the complex correlation structure inherent in contemporary labelled MS-based proteomics experiments by mixed models. However, unlike MSstatsTMT and msTrawler [@msTrawler2024], we avoid by default the imputation of missing peptide-spectrum match (PSM) intensities, instead assuming missingness at random after accounting for peptide effects. Nevertheless, users retain the option in msqrob2TMT to utilise the default imputation strategies available in QFeatures or to implement custom functions tailored to their specific needs. Additionally, DEqMS [@Zhu2020] and MSstatsTMT model the data upon summarisation at the protein level and msTrawler immediately models the data at the PSM level, while msqrob2TMT provides the user with the option to use either PSM-level or protein-level workflows. Another notable difference is that msqrob2 offers ridge regression and robust M-estimation, which can further stabilise parameter estimation and thus enhance reliable detection of differential proteins.

We demonstrate that our novel msqrob2TMT workflows outperform state-of-the-art tools by benchmarking them against DEqMS, MSstatsTMT, and msTrawler using two spike-in and real experimental dataset. Our comparisons also highlight that none of the state-of-the-art tools could provide valid inference across all of these three datasets, as their specific implementation limits the experimental designs for which they can infer DA. Particularly, MSstatsTMT cannot account for multiple covariates e.g., treatment effects together with block effects, age and/or other confounders, while DEqMS and msTrawler cannot address the complex correlation in TMT-experiments with multiple runs and technical replication, which are commonly employed in the proteomics community. Note, that Proteome Discoverer was not included in our benchmark, as we utilised datasets featured in the original MSstatsTMT publication, where the authors had already established the superior performance of MSstatsTMT compared to Proteome Discoverer [@Huang2020].

Importantly, our novel msqrob2TMT workflows offer researchers extensive flexibility to develop both peptide-level and protein-level data analysis workflows, incorporating custom preprocessing steps and user-defined models that can be tailored to their specific applications. The latest release further expands support for complex experimental designs by providing full compatibility with the model specification capabilities of the popular lme4 R package for mixed models, while enabling the use of both feature-level and sample-level covariates. The sample-level covariates allow msqrob2 to accommodate longitudinal and clustered experimental designs. The use of feature-level covariates is supported through integration with the QFeatures [@Qfeatures2022] data infrastructure. This integration ensures that raw input data are never lost but remain accessible and linked to the preprocessed, normalised, and summarised assays, as well as to the model outputs. This approach guarantees transparency, traceability, and reproducibility throughout the analysis workflow.

@@

The spike-in dataset was obtained from MassIVE (RMSV000000265) and has the following design: \sout{UPS1 peptides at concentrations of 500, 333, 250,  and 62.5 fmol}\textcolor{red}{500, 333, 250, and 62.5 fmol of UPS1 peptides} were spiked into 50 µg of SILAC HeLa peptides. This series forms a dilution gradient of 1, 0.667, 0.5, and 0.125 relative to the highest UPS1 peptide \sout{concentration}\textcolor{red}{amount} (500 fmol). A reference sample was prepared by combining the diluted UPS1 peptide samples \textcolor{red}{(286.5 fmol)} with 50 µg of SILAC HeLa peptides. Each dilution and the reference sample were processed in duplicate, yielding a total of ten samples. These samples were subsequently labelled using TMT10-plex reagents and combined for LC-MS/MS analysis, hereafter referred to as a "mixture." This protocol was repeated five times. Each mixture was analysed in technical triplicate, resulting in a total of 15 MS runs. This experimental design simulates a scenario with 10 biological replicates per condition, consisting of 2 replicates per condition within each of the five mixtures. \textcolor{red}{The MS data were searched by the MSstatsTMT authors using Proteome Discoverer 2.2.0.388 (Thermo Fisher Scientific) and Mascot Server 2.6.1 (Matrix Science, London, UK).}

@@

The msTrawler Multibatch Benchmarking Experiment data has been deposited on ProteomeXchange (PXD036799). However, for reproducibility of the results published in the original paper,
the data available on the google drive of the msTrawler paper (https://console.cloud.google.com/storage/browser/mstrawler_paper) was used. The experiment is a spike-in study which consists of a constant mouse plasma background in which yeast proteins were then spiked with the following dilution ratios 1, 2, 3, 5, 11, 14, 18, 20, 24, 28 and 32. The yeast proteins were harvested from different media; glycerol + ammonium sulfate, galactose + urea, galactose + monosodium glutamate, galactose + ammonium sulfate, glucose + ammonium sulfate, glucose + monosodium glutamate, and glucose + urea. The samples were labelled using the TMTpro 18-plex and pooled in six mixtures, each containing 15 samples. Two of the remaining TMT channels are reference channels, one reference channel is the combination of one batch and the other reference channel consists of all batches. For this dataset the ground truth is known: \textcolor{red}{yeast proteins were spiked in the samples in different amounts, while mouse proteins were added as a constant background. Hence, any yeast protein that is returned as significant is a true positive, and any mouse protein that is returned as significant is a false positive.} \sout{a true positive is a yeast protein that is returned as significant while a significant mouse protein is a false positive}.

The dataset imposes specific challenges for the normalisation. Indeed, the majority of the proteins are differentially abundant (DA), i.e. all yeast peptides, and they are all DA in the same direction, which implies a violation of the normalisation assumptions for most tools. To address this issue, the authors of msTrawler [@msTrawler2024] developed a custom normalisation workflow tailored specifically to this dataset. Particularly, they excluded mouse PSMs that correlate with the yeast dilution profile i.e., PSMs with a Pearson correlation coefficient higher than 0.25. Subsequently, PSMs with a summed signal-to-noise ratio below 20 were filtered out. Finally, the within sample geometric mean for the 50% most stable mouse PSMs in a run is used for normalisation. However, it should be noted that such a procedure is not applicable in typical experimental contexts, as the set of non-DA proteins is generally unknown. For this study, we nevertheless applied this custom msTrawler preprocessing workflow to the dataset prior to conducting differential analysis to ensure that all tools are benchmarked on the same preprocessed dataset, thereby enabling a fair comparison of their performance.

@@

The spike-in dataset was downloaded from MassIVE (RMSV000000264). In the study conducted by Plubell et al., 2017 [-@Plubell2017], twenty mice were divided into groups to investigate the effects of low-fat (LF) and high-fat (HF) diets. The experiment involves two factors: the type of diet (LF or HF), and the duration of the diet, categorised as either short-term (8 weeks) or long-term (18 weeks). This design resulted in four distinct groups corresponding to each combination of diet type and duration. Five mice, i.e. biological replicates, were randomly assigned to each group and samples from their epididymal adipose tissue were randomly assigned to three TMT10-plex mixtures. Within each mixture, two reference channels were included, each containing pooled samples representing a range of peptides from all samples. Not all TMT channels were utilised, leading to an unbalanced experimental design. Finally, each TMT mixture was fractionated into nine parts and analysed using synchronous precursor selection, culminating in a total of 27 MS runs.

@@

The DEqMS workflow [@Zhu2020] starts by $\log_2$-transforming the observed PSM intensities. The PSM intensities are then summarised into protein intensities by adopting the median sweep algorithm to the PSM data for each protein in each run, separately. This approach can be regarded as a one-step median polish \textcolor{red}{that corrects for PSM-specific effects (spectra effects). The PSM-specific effect is shared across all TMT intensities from the same spectrum, which can be estimated by their corresponding median. Hence, the median sweep algorithm corrects TMT intensities for PSM-specific effects by subtracting their corresponding median.} \sout{. Indeed, the $\log_2$-transformed PSM intensities are first centered by subtracting the corresponding median of the $\log_2$ PSM intensities from their corresponding spectrum the corresponding PSM.} Subsequently, the spectrum-centered PSM intensities in each TMT channel for each protein are aggregated into protein expression values using median summarisation. Finally, the protein summaries are normalised by subtracting the median of all summarised protein expression values within their corresponding channel.

Note, that for the msTrawler Multibatch Benchmarking Experiment, DEqMS starts from data that have been custom-filtered and normalised (see Section \ref{sec-multibatch}). Subsequently, the reference channels are removed, the data are $\log_2$ transformed, and summarised through a modified median sweep algorithm, which omits the normalisation step of the protein summaries.

@@

$$
\begin{array}{lcl}
y_{rc} &=& \mathbf{x}_{rc}^t \boldsymbol{\beta} + \beta_r^\text{run}  + \epsilon_{rc}\\
\epsilon_{rc}  &\sim& N(0,\sigma^2)
\end{array}
$$

with $y_{rc}$ the $\log_2$-transformed and summarised protein intensity for TMT reagent $c$ in MS-run $r$, $\mathbf{x}_{rc}^t$ the covariate pattern, e.g. diet, time, condition, dilution, and/or media for the sample in channel $c$ of run $r$ that is used to model the mean protein abundance upon correction for the run effect, $\boldsymbol{\beta}$ a vector of model parameters modelling the $\log_2$ fold changes (FC) for each covariate, and $\beta_r^\text{run}$ the fixed block effect for run r, and  $\epsilon_{rc}$ a normally distributed error term with mean 0 and variance $\sigma^2$ i.e., $N(0,\sigma^2)$. Hence, DEqMS can model data of complex experiments as long as these can be parameterised as a randomised complete block design. However, it cannot accomodate experiments with clustered or longitudinal designs, nor for designs including technical repeats.

Statistical hypothesis testing builds on the moderated t-tests and F-tests from the popular limma package. So users have the flexibility to infer DA using any linear combination of the model parameters ($\mathbf{L}^T\boldsymbol{\beta}$), which we also refer to as contrasts. DEqMS version 1.24 [@deqms] from Bioconductor release 3.20 is used in our manuscript.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{3}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}
\usepackage{anyfontsize}
\usepackage{multirow}
\usepackage{amsmath, booktabs, caption, longtable, xcolor, ulem}
\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={msqrob2TMT: robust linear mixed models for inferring differential abundant proteins in labelled experiments with arbitrarily complex design},
  pdfauthor={, , , , and },
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{msqrob2TMT: robust linear mixed models for inferring differential
abundant proteins in labelled experiments with arbitrarily complex
design}
\author{Stijn
Vandenbulcke\textsuperscript{$\dagger{}$,1,2,3} \and Christophe
Vanderaa\textsuperscript{$\dagger{}$,1} \and Oliver
Crook\textsuperscript{4} \and Lennart
Martens\textsuperscript{2,3} \and Lieven Clement\textsuperscript{1,*}}
\date{}

\begin{document}
\maketitle

\textsuperscript{$\dagger{}$}
These authors contributed equally to this work.

\textsuperscript{1} Department of Mathematics, Computer Science and
Statistics, Ghent University, Ghent, Belgium\\
\textsuperscript{2} CompOmics, VIB Center for Medical Biotechnology,
VIB, Ghent, Belgium\\
\textsuperscript{3} Department of Biomolecular Medicine, Faculty of
Medicine and Health Sciences, Ghent University, Ghent, Belgium\\
\textsuperscript{4} Department of Statistics, University of Oxford,
Oxford, UK.

\textsuperscript{*} Correspondence:
\href{mailto:lieven.clement@ugent.be}{Lieven Clement
\textless{}lieven.clement@ugent.be\textgreater{}}

@@

@@

High-throughput LC-MS-based proteomic workflows are widely used to
quantify differential protein abundance across samples. Relative protein
quantification is generally achieved through either label-free or
labelled workflows. The latter employ stable isotope labelling
techniques such as metabolic and post-metabolic labelling, which gained
traction in recent years (Li et al. 2021). Compared to label-free
approaches, labelling offers the advantage of mitigating run-to-run
variability in peptide identification and quantification by pooling and
analysing multiple samples within a single run. This enables researchers
to compare proteomes across multiple conditions or treatments within a
single MS run, thereby providing a more comprehensive view of the
proteome and enhancing the statistical power of the analysis. Current
tandem mass tag (TMT) kits, for instance, support the multiplexing of up
to 18 samples in a single run (Li et al. 2021).

However, contemporary experiments often involve increasingly complex
designs, where the number of samples exceeds the capacity of a single
run. This has far reaching consequences for the downstream data
analysis. In labelled experiments with multiple MS runs, biological
replicates are often distributed across distinct runs, and additional
technical MS runs are frequently included. Similar to label-free
approaches, such multi-run experiments are prone to numerous missing
peptide intensity values across MS runs (Brenes et al. 2019).
Furthermore, the data of these designs are correlated at multiple
levels. Indeed, the MS-technology introduces various sources of
technical variation, including run-level effects, channel effects, and
spectrum-specific effects, amongst others. Hence, ion intensities from
samples multiplexed in the same run are more similar than those from
samples multiplexed in different runs. Additionally, quantification does
not happen at the protein level but at the level of its identified
peptide ions. Peptide ion abundances from the same sample within a run
are inherently more similar than those of peptide ions from different
samples. At the lowest level, TMT reporter ions from a specific spectrum
enable peptide ion quantification across multiple samples, resulting in
correlated ion intensities; ion intensities within the same spectrum are
more alike than those measured across different spectra. Furthermore,
mixtures of multiplexed samples are often run in technical repeats on
the MS, introducing yet another level of correlation. Addressing such a
complex correlation structure in labelled experiments necessitates
flexible statistical models capable of producing valid tests for
differential abundance. However, most proteomics software tools assume
independence among the (summarised) ion intensities and can therefore
not account for the abovementioned intricate correlation structures in
contemporary TMT experiments, which may in turn lead to an inflation of
false positives (Huang et al. 2020b). The MSstatsTMT tool (Huang et al.
2020b), however, is one of the notable exceptions, as it has been
explicitly developed to accommodate experiments with multiple
conditions, biological replicate runs, technical repeat runs, and
unbalanced designs.

In this publication, we present a bespoke data analysis workflow for
labelled MS-based proteomics experiments, embedded in our msqrob2
universe in R/Bioconductor. Initially designed for label-free workflows,
msqrob2 has demonstrated advantages over other state-of-the-art tools in
terms of both performance and its modular, transparent implementation
(Sticker et al. 2020). In this work, we extend these benefits to the
analysis of labelled experiments. Similar to MSstatsTMT, our workflows
address the complex correlation structure inherent in contemporary
labelled MS-based proteomics experiments by mixed models. However,
unlike MSstatsTMT and msTrawler (O'Brien et al. 2024), we avoid by
default the imputation of missing peptide-spectrum match (PSM)
intensities, instead assuming missingness at random after accounting for
peptide effects. Nevertheless, users retain the option in msqrob2TMT to
utilise the default imputation strategies available in QFeatures or to
implement custom functions tailored to their specific needs.
Additionally, DEqMS (Zhu et al. 2020) and MSstatsTMT model the data upon
summarisation at the protein level and msTrawler immediately models the
data at the PSM level, while msqrob2TMT provides the user with the
option to use either PSM-level or protein-level workflows. Another
notable difference is that msqrob2 offers ridge regression and robust
M-estimation, which can further stabilise parameter estimation and thus
enhance reliable detection of differential proteins.

We demonstrate that our novel msqrob2TMT workflows outperform
state-of-the-art tools by benchmarking them against DEqMS, MSstatsTMT,
and msTrawler using two spike-in and real experimental dataset. Our
comparisons also highlight that none of the state-of-the-art tools could
provide valid inference across all of these three datasets, as their
specific implementation limits the experimental designs for which they
can infer DA. Particularly, MSstatsTMT cannot account for multiple
covariates e.g., treatment effects together with block effects, age
and/or other confounders, while DEqMS and msTrawler cannot address the
complex correlation in TMT-experiments with multiple runs and technical
replication, which are commonly employed in the proteomics community.
Note, that Proteome Discoverer was not included in our benchmark, as we
utilised datasets featured in the original MSstatsTMT publication, where
the authors had already established the superior performance of
MSstatsTMT compared to Proteome Discoverer (Huang et al. 2020b).

Importantly, our novel msqrob2TMT workflows offer researchers extensive
flexibility to develop both peptide-level and protein-level data
analysis workflows, incorporating custom preprocessing steps and
user-defined models that can be tailored to their specific applications.
The latest release further expands support for complex experimental
designs by providing full compatibility with the model specification
capabilities of the popular lme4 R package for mixed models, while
enabling the use of both feature-level and sample-level covariates. The
sample-level covariates allow msqrob2 to accommodate longitudinal and
clustered experimental designs. The use of feature-level covariates is
supported through integration with the QFeatures (Gatto and Vanderaa
2022) data infrastructure. This integration ensures that raw input data
are never lost but remain accessible and linked to the preprocessed,
normalised, and summarised assays, as well as to the model outputs. This
approach guarantees transparency, traceability, and reproducibility
throughout the analysis workflow.

@@

The spike-in dataset was obtained from MassIVE (RMSV000000265) and has
the following design:
\sout{UPS1 peptides at concentrations of 500, 333, 250,  and 62.5 fmol}\textcolor{red}{500, 333, 250, and 62.5 fmol of UPS1 peptides}
were spiked into 50 µg of SILAC HeLa peptides. This series forms a
dilution gradient of 1, 0.667, 0.5, and 0.125 relative to the highest
UPS1 peptide \sout{concentration}\textcolor{red}{amount} (500 fmol). A
reference sample was prepared by combining the diluted UPS1 peptide
samples \textcolor{red}{(286.5 fmol)} with 50 µg of SILAC HeLa peptides.
Each dilution and the reference sample were processed in duplicate,
yielding a total of ten samples. These samples were subsequently
labelled using TMT10-plex reagents and combined for LC-MS/MS analysis,
hereafter referred to as a ``mixture.'' This protocol was repeated five
times. Each mixture was analysed in technical triplicate, resulting in a
total of 15 MS runs. This experimental design simulates a scenario with
10 biological replicates per condition, consisting of 2 replicates per
condition within each of the five mixtures.
\textcolor{red}{The MS data were searched by the MSstatsTMT authors using Proteome Discoverer 2.2.0.388 (Thermo Fisher Scientific) and Mascot Server 2.6.1 (Matrix Science, London, UK).}

@@

The msTrawler Multibatch Benchmarking Experiment data has been deposited
on ProteomeXchange (PXD036799). However, for reproducibility of the
results published in the original paper, the data available on the
google drive of the msTrawler paper
(https://console.cloud.google.com/storage/browser/mstrawler\_paper) was
used. The experiment is a spike-in study which consists of a constant
mouse plasma background in which yeast proteins were then spiked with
the following dilution ratios 1, 2, 3, 5, 11, 14, 18, 20, 24, 28 and 32.
The yeast proteins were harvested from different media; glycerol +
ammonium sulfate, galactose + urea, galactose + monosodium glutamate,
galactose + ammonium sulfate, glucose + ammonium sulfate, glucose +
monosodium glutamate, and glucose + urea. The samples were labelled
using the TMTpro 18-plex and pooled in six mixtures, each containing 15
samples. Two of the remaining TMT channels are reference channels, one
reference channel is the combination of one batch and the other
reference channel consists of all batches. For this dataset the ground
truth is known:
\textcolor{red}{yeast proteins were spiked in the samples in different amounts, while mouse proteins were added as a constant background. Hence, any yeast protein that is returned as significant is a true positive, and any mouse protein that is returned as significant is a false positive.}
\sout{a true positive is a yeast protein that is returned as significant while a significant mouse protein is a false positive}.

The dataset imposes specific challenges for the normalisation. Indeed,
the majority of the proteins are differentially abundant (DA), i.e.~all
yeast peptides, and they are all DA in the same direction, which implies
a violation of the normalisation assumptions for most tools. To address
this issue, the authors of msTrawler (O'Brien et al. 2024) developed a
custom normalisation workflow tailored specifically to this dataset.
Particularly, they excluded mouse PSMs that correlate with the yeast
dilution profile i.e., PSMs with a Pearson correlation coefficient
higher than 0.25. Subsequently, PSMs with a summed signal-to-noise ratio
below 20 were filtered out. Finally, the within sample geometric mean
for the 50\% most stable mouse PSMs in a run is used for normalisation.
However, it should be noted that such a procedure is not applicable in
typical experimental contexts, as the set of non-DA proteins is
generally unknown. For this study, we nevertheless applied this custom
msTrawler preprocessing workflow to the dataset prior to conducting
differential analysis to ensure that all tools are benchmarked on the
same preprocessed dataset, thereby enabling a fair comparison of their
performance.

@@

The spike-in dataset was downloaded from MassIVE (RMSV000000264). In the
study conducted by Plubell et al., 2017 (2017), twenty mice were divided
into groups to investigate the effects of low-fat (LF) and high-fat (HF)
diets. The experiment involves two factors: the type of diet (LF or HF),
and the duration of the diet, categorised as either short-term (8 weeks)
or long-term (18 weeks). This design resulted in four distinct groups
corresponding to each combination of diet type and duration. Five mice,
i.e.~biological replicates, were randomly assigned to each group and
samples from their epididymal adipose tissue were randomly assigned to
three TMT10-plex mixtures. Within each mixture, two reference channels
were included, each containing pooled samples representing a range of
peptides from all samples. Not all TMT channels were utilised, leading
to an unbalanced experimental design. Finally, each TMT mixture was
fractionated into nine parts and analysed using synchronous precursor
selection, culminating in a total of 27 MS runs.

@@

The DEqMS workflow (Zhu et al. 2020) starts by \(\log_2\)-transforming
the observed PSM intensities. The PSM intensities are then summarised
into protein intensities by adopting the median sweep algorithm to the
PSM data for each protein in each run, separately. This approach can be
regarded as a one-step median polish
\textcolor{red}{that corrects for PSM-specific effects (spectra effects). The PSM-specific effect is shared across all TMT intensities from the same spectrum, which can be estimated by their corresponding median. Hence, the median sweep algorithm corrects TMT intensities for PSM-specific effects by subtracting their corresponding median.}
\sout{. Indeed, the $\log_2$-transformed PSM intensities are first centered by subtracting the corresponding median of the $\log_2$ PSM intensities from their corresponding spectrum the corresponding PSM.}
Subsequently, the spectrum-centered PSM intensities in each TMT channel
for each protein are aggregated into protein expression values using
median summarisation. Finally, the protein summaries are normalised by
subtracting the median of all summarised protein expression values
within their corresponding channel.

Note, that for the msTrawler Multibatch Benchmarking Experiment, DEqMS
starts from data that have been custom-filtered and normalised (see
Section \ref{sec-multibatch}). Subsequently, the reference channels are
removed, the data are \(\log_2\) transformed, and summarised through a
modified median sweep algorithm, which omits the normalisation step of
the protein summaries.

@@

\[
\begin{array}{lcl}
y_{rc} &=& \mathbf{x}_{rc}^t \boldsymbol{\beta} + \beta_r^\text{run}  + \epsilon_{rc}\\
\epsilon_{rc}  &\sim& N(0,\sigma^2)
\end{array}
\]

with \(y_{rc}\) the \(\log_2\)-transformed and summarised protein
intensity for TMT reagent \(c\) in MS-run \(r\), \(\mathbf{x}_{rc}^t\)
the covariate pattern, e.g.~diet, time, condition, dilution, and/or
media for the sample in channel \(c\) of run \(r\) that is used to model
the mean protein abundance upon correction for the run effect,
\(\boldsymbol{\beta}\) a vector of model parameters modelling the
\(\log_2\) fold changes (FC) for each covariate, and
\(\beta_r^\text{run}\) the fixed block effect for run r, and
\(\epsilon_{rc}\) a normally distributed error term with mean 0 and
variance \(\sigma^2\) i.e., \(N(0,\sigma^2)\). Hence, DEqMS can model
data of complex experiments as long as these can be parameterised as a
randomised complete block design. However, it cannot accomodate
experiments with clustered or longitudinal designs, nor for designs
including technical repeats.

Statistical hypothesis testing builds on the moderated t-tests and
F-tests from the popular limma package. So users have the flexibility to
infer DA using any linear combination of the model parameters
(\(\mathbf{L}^T\boldsymbol{\beta}\)), which we also refer to as
contrasts. DEqMS version 1.24 (Zhu 2023) from Bioconductor release 3.20
is used in our manuscript.
